Hydra Battles and Cie

### 2021-09-08 doc/chapter-powers.tex
adapt chapter-powers to Alectryon (partial)

\begin{Coqsrc}
Fixpoint binary_power_mult (x a:A)(p:positive) : A 
  :=
  match p with
    | xH =>  a * x
    | xO q => binary_power_mult  (x * x) a q
    | xI q =>  binary_power_mult  (x * x) (a * x) q
  end.

Fixpoint Pos_bpow (x:A)(p:positive) :=
 match p with
  | xH => x
  | xO q => Pos_bpow  (x * x) q
  | xI q => binary_power_mult   (x * x) x q
end.

Definition N_bpow x (n:N) := 
  match n with 
  | 0%N => 1
  | Npos p => Pos_bpow x p
  end.

End Definitions.
\end{Coqsrc}

Let us  close the section \texttt{Definitions} and mark the argument \texttt{A} as implicit.

\begin{Coqsrc}
End Definitions.

Arguments N_bpow {A}.
Arguments power {A}.
\end{Coqsrc}

@@

\subsubsection{Exponentiation on $2\times 2$ matrices}
\label{naive-matrix}
Our second example is a definition of $M^n$ where $M$ is a $2\times 2$ matrix
over any ``scalar''  type $A$, assuming one can provide $A$ with a semi-ring structure~\cite{Coq}.

%\subsubsection{Representation of $2\times 2$ matrices}

A $2\times 2$ matrix will be simply represented by a structure with four fields;
each field \texttt{c$ij$} is associated with the $i$-th line and $j$-th column of the considered matrix.

\inputsnippets{FirstSteps/M2a}

@@

The structure type \texttt{M2.t} allows us to define the product
of two matrices.

\inputsnippets{FirstSteps/M2Mult}

@@

We define two monoids on type \texttt{nat}:
\begin{itemize}
\item The ``natural'' monoid $(\mathbb{N},\times, 1)$ :

  \begin{Coqsrc}
Instance nat_mult_op : Mult_op nat | 5 := Nat.mul.

Instance  Natmult : Monoid nat_mult_op  1%nat | 5
Proof.
   split;unfold nat_mult_op, mult_op; intros; ring.
Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Instance nat_plus_op : Mult_op nat | 12 := Nat.add.

Instance Natplus : Monoid nat_plus_op  0%nat | 12.
(* Proof omitted *)
\end{Coqsrc}
\end{itemize}

@@

\begin{Coqsrc}
Instance int31_mult_op : Mult_op int31 := mul31.

Instance  Int31mult : Monoid int31_mult_op  1.
Proof.
   split;unfold int31_mult_op, mult_op; intros; ring.
Qed.
\end{Coqsrc}

@@

Fixpoint int31_from_nat (n:nat) :=
  match n with
  | O => 1
  | S p => 1 + int31_from_nat p
  end.

Coercion int31_from_nat : nat >-> int31.

Fixpoint fact (n:nat) := 
  match n with
   | O => 1
   | S p => n * fact p
  end.

Example fact_zero : exists n:nat, fact n = 0.
Proof.  now exists 40%nat.  Qed.

End Bad.
\end{Coqbad}

@@

In Sect.\vref{naive-matrix}, we defined a function for computing powers of any $2\times 2$ 
matrix over any semi-ring. For proving a simple property of matrix exponentiation, we had
 to prove that matrix multiplication is associative and admits the identity matrix as a neutral element. These properties are easily expressed within the type class framework, by defining a \emph{family} of monoids.
It suffices to define an instance of \texttt{Monoid} within the scope of a hypothesis
of type \texttt{semi\_ring\_theory}.

\begin{Coqsrc}
Section M2_def.
Variables (A:Type)
           (zero one : A) 
           (plus mult  : A -> A -> A).

 Variable rt : semi_ring_theory  zero one plus mult  (@eq A).
 Add  Ring Aring : rt.
\end{Coqsrc}

\begin{Coqsrc}
Structure M2 : Type := {c00 : A;  c01 : A;
                        c10 : A;  c11 : A}.

Definition Id2 : M2 := Build_M2 1 0 0 1.

Definition M2_mult (m m':M2) : M2 :=
 Build_M2 
          (c00 m * c00 m' + c01 m * c10 m')
          (c00 m * c01 m' + c01 m * c11 m')
          (c10 m * c00 m' + c11 m * c10 m')
          (c10 m * c01 m' + c11 m * c11 m').

Global Instance M2_op : Mult_op M2 := M2_mult.
\end{Coqsrc}

\begin{Coqsrc}
Global Instance M2_Monoid : Monoid   M2_op Id2.
(* Proof omitted *)

End M2_def.

Arguments M2_Monoid {A zero one plus mult} rt.
\end{Coqsrc}

@@

\begin{Coqsrc}
Class Equiv A := equiv : relation A.

Infix "==" := equiv (at level 70) : type_scope.
\end{Coqsrc}

@@

\begin{Coqsrc}
Class EMonoid (A:Type)(E_op : Mult_op A)(E_one : A) 
      (E_eq: Equiv A): Prop :=
  {
    Eq_equiv :> Equivalence equiv;
    Eop_proper :> Proper (equiv ==> equiv ==> equiv) E_op;
    Eop_assoc : forall x y z, x * (y * z) == x * y * z;
    Eone_left : forall x,  E_one * x == x;
    Eone_right : forall x,  x * E_one ==  x
  }.
\end{Coqsrc}

@@

\begin{Coqsrc}
Global Instance eq_equiv {A} : Equiv A := eq.

Global Instance Monoid_EMonoid `(M:@Monoid A op one) :
        EMonoid  op one eq_equiv.
Proof.
split; unfold eq_equiv, equiv in *.
 - apply eq_equivalence.
 - intros x y H z t H0; now subst.
 - intros; now rewrite (op_assoc).
 - intro; now rewrite one_left.
 - intro; now rewrite one_right.
Defined.
\end{Coqsrc}

@@

\begin{Coqsrc}
Coercion Monoid_EMonoid : Monoid >-> EMonoid.
\end{Coqsrc}

%ici%

@@

\begin{Coqsrc}
Check NMult : EMonoid  N.mul 1%N eq.
\end{Coqsrc}

\begin{Coqanswer}
  NMult:EMonoid N.mul 1%N eq
     : EMonoid N.mul 1%N eq
\end{Coqanswer}

>>>>

\inputsnippets{FirstSteps/bpowDef}
Let us  close the section \texttt{Definitions} and mark the argument \texttt{A} as implicit.
\inputsnippets{FirstSteps/EndDefs}

@@

% \subsubsection{Exponentiation on $2\times 2$ matrices}
% \label{naive-matrix}
% Our second example is a definition of $M^n$ where $M$ is a $2\times 2$ matrix
% over any ``scalar''  type $A$, assuming one can provide $A$ with a semi-ring structure~\cite{Coq}.

% %\subsubsection{Representation of $2\times 2$ matrices}

% A $2\times 2$ matrix will be simply represented by a structure with four fields;
% each field \texttt{c$ij$} is associated with the $i$-th line and $j$-th column of the considered matrix.

@@

@@

We define two monoids on type \texttt{nat}:
\begin{itemize}
\item The ``natural'' monoid $(\mathbb{N},\times, 1)$ :
\inputsnippets{Monoid_instances/natMult}

@@

\inputsnippets{Monoid_instances/natPlus}
\end{itemize}

@@

\inputsnippets{Monoid_instances/int31}

@@

\inputsnippets{Monoid_instances/BadFact}

@@

Let $(A,+,\times)$ be a semi-ring. We define a multiplicative monoid on the set of \emph{e.g.} $2\times 2$-) square matrices over $A$.
It suffices to define an instance of \texttt{Monoid} within the scope of a hypothesis
of type \texttt{semi\_ring\_theory}.

\inputsnippets{Monoid_instances/M2Defsa, Monoid_instances/M2Defsb}

@@

\inputsnippets{Monoid_def/EquivDef}

@@

\inputsnippets{Monoid_def/EMonoidDef}

@@

\inputsnippets{Monoid_def/Coerciona}

@@

\inputsnippets{Monoid_def/Coercionb}

@@

\inputsnippets{Monoid_instances/CheckCoercion}
